---
title: "Introduction to the SISE Package"
author: "Justin D. Tubbs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "svg")
```

## Smoothed Interval-based Survival Estimation

The **SISE** package is used to estimate smoothed survival (and time-to-event) distributions from interval-censored data, building on the traditional Turnbull (1976) estimator. In a similar manner, our method can also be used when observations are exact or right-censored using a modification of the Kaplan-Meier (1958) procedure.

This vignette utilizes a dataset on breast cosmesis (Finkelstein & Wolfe,1985) loaded from the `interval` R package.

```{r}
library(SISE)

data(bcos, package = "interval")
```


The main function from the **SISE** package is the `smoothTB()` function. It takes a data frame or matrix of interval-censored observations as input, where columns 1 and 2 are the left and right interval bounds observed for each subject. For left-censored observations, their value in column 1 should be `NA`. Likewise, right-censored observations should have `NA` in column 2.

Here we will re-format the breast cosmesis data so that it can be passed to the `smoothTB()` function.

```{r}
dat <- bcos[,1:2]
dat$left[dat$left == 0] <- NA
dat$right[is.infinite(dat$right)] <- NA
```

Now we can provide this data frame to the `smoothTB()` function. Although it is not known from the dataset description, we assume here that the average number of observations per subject, `n.obs` is 5. We assume the left boundary, `left.bound` of event times is 0. We will use the `logNe` penalty for choosing a bandwidth. Further, we need to specify the number of decimal points of the observed data. We leave the `tolerance` and `inflection.threshold` at their default values.

To estimate survival curves for the two different treatment groups, we run the function on the two groups independently.

```{r}
results1 <- smoothTB(dat = dat[bcos$treatment == "Rad",], 
                     n.obs = 5, 
                     left.bound = 0, 
                     penalty=c('logNe'), 
                     n.dec = 0)
results2 <- smoothTB(dat = dat[bcos$treatment == "RadChem",],
                     n.obs = 5, 
                     left.bound = 0, 
                     penalty=c('logNe'), 
                     n.dec = 0)
str(results1)
```

The output of the `smoothTB()` function is a list object holding intermediate variables used in the analysis, some characteristics of the sample, and importantly, the estimated smooth survival and onset distributions, called `TB.s.smooth.xxx` and `TB.o.smooth.xxx` respectively, where *xxx* is the sBIC penalty parameter used when choosing the smoothing bandwidth.

Thus, we can plot the survival curve estimated by our method and compare it to that estimated using the classic Turnbull algorithm.

```{r, fig.width=7}
library(ggplot2)

plt.df <- data.frame(Time = c(results1$time.s, results2$time.s),
                     Survival = c(results1$TB.s.smooth.logNe,
                                  results2$TB.s.smooth.logNe),
                     Treatment = factor(c(rep("Rad", length(results1$TB.s.smooth.logNe)),
                                          rep("RadChem", length(results2$TB.s.smooth.logNe)))))

p1 <- ggplot(plt.df, aes(x = Time, y = Survival, color = Treatment)) +
  geom_step() + theme_bw() + 
  labs(title = " Smoothed Turnbull Estimate", x = "Time (months)") + 
  theme(legend.position = "bottom")

plt.df.tb <- data.frame(Time = c(results1$time.s, results2$time.s),
                     Survival = c(results1$TB.s, results2$TB.s),
                     Treatment = factor(c(rep("Rad", length(results1$TB.s)),
                                          rep("RadChem", length(results2$TB.s)))))

p2 <- ggplot(plt.df.tb, aes(x = Time, y = Survival, color = Treatment)) +
  geom_step() + theme_bw() + 
  labs(title = "Turnbull Estimate", x = "Time (months)") + 
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p1,p2, nrow=1)

```


The **SISE** package also provides a helpful function, `impute.time()`, for imputing censored observations given an assumed onset distribution estimate. To demonstrate, we predict onsets of left- and interval-censored individuals in the two groups, given the estimated smoothed onset curves. 


```{r, fig.width=5}
pred1 <- impute.time(dat = results1$dat, 
                       event.distribution = results1$TB.o.smooth.logNe, 
                       time = results1$time.o, 
                       type = "LI", n.dec = 0)
pred2 <- impute.time(dat = results2$dat, 
                       event.distribution = results2$TB.o.smooth.logNe, 
                       time = results2$time.o, 
                       type = "LI", n.dec = 0)

plt.pred <- data.frame(onset = c(pred1, pred2),
                       Treatment = c(rep("Rad", length(pred1)),
                                     rep("RadChem", length(pred2))))

ggplot(plt.pred, aes(x = Treatment, y = onset, fill = Treatment)) +
  geom_violin() +
  geom_boxplot(fill = "white", alpha = 0.7) +
  theme_bw() + labs(title = "Predicted Onset Time Across Treatment Groups",
                    y = "Time to Onset (months)")

```
